---
title: "*Spatial Statistics Lab 7*"
author: '**Onyedikachi J Okeke**'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **0.0 To Load the library**
```{r,comment="",echo=TRUE,eval=FALSE}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
library(rgdal)
library(rgdal)
library(terra)
library(RColorBrewer)
library(classInt)
library(epitools)
library(DCluster)

```
##### Library
```{r,echo=F,results='Show', fig.show='hold',eval=T}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
library(rgdal)
library(rgdal)
library(terra)
library(RColorBrewer)
library(classInt)
library(epitools)
library(DCluster)

```

### **Q1 Loading Sids data**
```{r,comment="",echo=TRUE,eval=FALSE}
data(nc.sids)
sids<-data.frame(Observed=nc.sids$SID74)
#combine the expected number of SID cases to the data frame
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
#combine x, y coordinates to the data frame
sids<-cbind(sids, x=nc.sids$x, y=nc.sids$y)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
data(nc.sids)
sids<-data.frame(Observed=nc.sids$SID74)
#combine the expected number of SID cases to the data frame
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
#combine x, y coordinates to the data frame
sids<-cbind(sids, x=nc.sids$x, y=nc.sids$y)

```

### **GAM using the centroids of the areas in data**
```{r,comment="",echo=TRUE,eval=FALSE}
#step is the step of the grid, alpha is the Significance level of the tests performed.
sidsgam<-opgam(data=sids, radius=30, step=10, alpha=.002)

#Plot centroids
plot(sids$x, sids$y, xlab="Easting", ylab="Northing")
#Plot points marked as clusters
points(sidsgam$x, sidsgam$y, col="red", pch="*")
title("Alpha level = 0.002")

#Change the alpha (Significance level of the tests performed) to 0.05 and rerun the GAM
sidsgam05<-opgam(data=sids, radius=30, step=10, alpha=.05)

#Plot centroids
plot(sids$x, sids$y, xlab="Easting", ylab="Northing")
#Plot points marked as clusters
points(sidsgam05$x, sidsgam05$y, col="red", pch="*")
title("Alpha level = 0.05")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
#step is the step of the grid, alpha is the Significance level of the tests performed.
sidsgam<-opgam(data=sids, radius=30, step=10, alpha=.002)

#Plot centroids
plot(sids$x, sids$y, xlab="Easting", ylab="Northing")
#Plot points marked as clusters
points(sidsgam$x, sidsgam$y, col="red", pch="*")
title("Alpha level = 0.002")

#Change the alpha (Significance level of the tests performed) to 0.05 and rerun the GAM
sidsgam05<-opgam(data=sids, radius=30, step=10, alpha=.05)

#Plot centroids
plot(sids$x, sids$y, xlab="Easting", ylab="Northing")
#Plot points marked as clusters
points(sidsgam05$x, sidsgam05$y, col="red", pch="*")
title("Alpha level = 0.05")

```

Q1. Describe the pattern
Q2. Change the alpha (Significance level of the tests performed) to 0.05 and rerun the GAM. Describe the differences between it and pattern from Q1.

### **Kulldorff’s spatial scan statistics**
```{r,comment="",echo=TRUE,eval=FALSE}
sids<-data.frame(Observed=nc.sids$SID74)
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
#combine risk population, and x, y coordinates to the data frame
sids<-cbind(sids, Population=nc.sids$BIR74, x=nc.sids$x, y=nc.sids$y)
# Kulldorff’s spatial scan statistics method over the centroids
# there are different models and here we use negative binomial
mle<-calculate.mle(sids, model="negbin") # Calculate parameters in sampling procedures

knresults<-opgam(data=sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=kn.iscluster, fractpop=.15, R=99, model="negbin", mle=mle)
#Plot all centroids and significant ones in red

plot(sids$x, sids$y, main="Kulldorff and Nagarwalla's method, Fractpop = 0.15")
points(knresults$x, knresults$y, col="red", pch=19)

#Plot the cluster with the highest likelihood ratio test in green
clusters<-get.knclusters(sids, knresults)
idx<-which.max(knresults$statistic)
points(sids$x[clusters[[idx]]], sids$y[clusters[[idx]]], col="green", pch=19)


#Change the maximum fraction of the total population to 0.5 and rerun the test
knresults05<-opgam(data=sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=kn.iscluster, fractpop=.5, R=99, model="negbin", mle=mle)

#Plot all centroids and significant ones in red
plot(sids$x, sids$y, main="Kulldorff and Nagarwalla's method, Fractpop = 0.5")
points(knresults05$x, knresults05$y, col="red", pch=19)

#Plot the cluster with the highest likelihood ratio test in green
clusters<-get.knclusters(sids, knresults05)
idx<-which.max(knresults05$statistic)
points(sids$x[clusters[[idx]]], sids$y[clusters[[idx]]], col="green", pch=19)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
sids<-data.frame(Observed=nc.sids$SID74)
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
#combine risk population, and x, y coordinates to the data frame
sids<-cbind(sids, Population=nc.sids$BIR74, x=nc.sids$x, y=nc.sids$y)
# Kulldorff’s spatial scan statistics method over the centroids
# there are different models and here we use negative binomial
mle<-calculate.mle(sids, model="negbin") # Calculate parameters in sampling procedures

knresults<-opgam(data=sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=kn.iscluster, fractpop=.15, R=99, model="negbin", mle=mle)
#Plot all centroids and significant ones in red

plot(sids$x, sids$y, main="Kulldorff and Nagarwalla's method, Fractpop = 0.15")
points(knresults$x, knresults$y, col="red", pch=19)

#Plot the cluster with the highest likelihood ratio test in green
clusters<-get.knclusters(sids, knresults)
idx<-which.max(knresults$statistic)
points(sids$x[clusters[[idx]]], sids$y[clusters[[idx]]], col="green", pch=19)


#Change the maximum fraction of the total population to 0.5 and rerun the test
knresults05<-opgam(data=sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=kn.iscluster, fractpop=.5, R=99, model="negbin", mle=mle)

#Plot all centroids and significant ones in red
plot(sids$x, sids$y, main="Kulldorff and Nagarwalla's method, Fractpop = 0.5")
points(knresults05$x, knresults05$y, col="red", pch=19)

#Plot the cluster with the highest likelihood ratio test in green
clusters<-get.knclusters(sids, knresults05)
idx<-which.max(knresults05$statistic)
points(sids$x[clusters[[idx]]], sids$y[clusters[[idx]]], col="green", pch=19)


```

Q3. Describe the pattern. 
Q4. Change the maximum fraction of the total population to 0.5 and rerun the test. Take a screenshot of the results. Describe the differences between it and pattern from Q3. (15 points)

### **Besag and Newell's Statistic for Spatial Clustering**
```{r,comment="",echo=TRUE,eval=FALSE}
sids<-data.frame(Observed=nc.sids$SID74)
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
sids<-cbind(sids, x=nc.sids$x, y=nc.sids$y)

#Where K = 10
bnresults10<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=10, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults10$x, bnresults10$y, col="red", pch=19)
title("Besag and Newell’s test, K = 10")

#Where K = 20
bnresults20<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=20, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults20$x, bnresults20$y, col="red", pch=19)
title("Besag and Newell’s test, K = 20")

#Where K = 30
bnresults30<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=30, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults30$x, bnresults30$y, col="red", pch=19)
title("Besag and Newell’s test, K = 30")

#Where K = 40
bnresults40<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=40, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults40$x, bnresults40$y, col="red", pch=19)
title("Besag and Newell’s test, K = 40")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
sids<-data.frame(Observed=nc.sids$SID74)
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
sids<-cbind(sids, x=nc.sids$x, y=nc.sids$y)

#Where K = 10
bnresults10<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=10, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults10$x, bnresults10$y, col="red", pch=19)
title("Besag and Newell’s test, K = 10")

#Where K = 20
bnresults20<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=20, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults20$x, bnresults20$y, col="red", pch=19)
title("Besag and Newell’s test, K = 20")

#Where K = 30
bnresults30<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=30, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults30$x, bnresults30$y, col="red", pch=19)
title("Besag and Newell’s test, K = 30")

#Where K = 40
bnresults40<-opgam(sids, thegrid=sids[,c("x","y")], alpha=.05,
iscluster=bn.iscluster, set.idxorder=TRUE, k=40, model="poisson",
R=100, mle=calculate.mle(sids) )
#Plot all the centroids
plot(sids$x, sids$y)
#Plot significant centroids in red
points(bnresults40$x, bnresults40$y, col="red", pch=19)
title("Besag and Newell’s test, K = 40")


```

Q5. Take a screenshot of the results. Describe the pattern. 
Q6. Change the k (e.g. 10, 30) and rerun the Besag and Newell’s test. Take a screen shot of each plot. Compare these patterns against that from Q3.

### **Stone test**
```{r,comment="",echo=TRUE,eval=FALSE}
sids<-data.frame(Observed=nc.sids$SID74)
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
sids<-cbind(sids, x=nc.sids$x, y=nc.sids$y)
#Compute Stone's statistic around a county
region<-which(row.names(nc.sids)=="Robeson")
stone.stat(sids, region=region, lambda=1)
stone.test(Observed~offset(log(Expected)), sids, model="poisson", R=99,
region=region, lambda=1)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
sids<-data.frame(Observed=nc.sids$SID74)
sids<-cbind(sids, Expected=nc.sids$BIR74*sum(nc.sids$SID74)/sum(nc.sids$BIR74))
sids<-cbind(sids, x=nc.sids$x, y=nc.sids$y)
#Compute Stone's statistic around a county
region<-which(row.names(nc.sids)=="Robeson")
stone.stat(sids, region=region, lambda=1)
stone.test(Observed~offset(log(Expected)), sids, model="poisson", R=99,
region=region, lambda=1)

```

Q7. What is the pattern of this county? 
Q8. Compare results generated from the above three methods and those from local Getis G* and local moran’s I (in Lab 6). Explain the differences if any.